Hydras \& Co.: Formalized mathematics in Coq\\ for inspiration and entertainment

### 2022-06-24 doc/JFLA2022/Slides/Slides.tex
last changes in Slides ?

\section{Kirby \& Paris' hydras}

\begin{frame}
  \frametitle{Kirby \& Paris' hydras}
 \begin{figure}[h]
    \centering
    \begin{tikzpicture}[very thick, scale=0.20]
      \node (h1) at (4,0){$\bullet$};
      \node[blue] (h2) at (4,2){$\bullet$};
      \node[blue] (h3) at (2,4){$\bullet$};
      \node[blue] (h4) at (0,6){$\bullet$};
      \node[red!90!black] (h7) at (6,2){$\bullet$};
      \node[blue] (h5) at (4,6){$\bullet$};
      \node[blue] (h6) at (6,4){$\bullet$};
      \draw (h1) -- (h2) ;
      \draw (h2) -- (h3) ;
      \draw (h1) -- node[red!90!black,font=\small,sloped,
      shift={(0.01,-0.075)},rotate=90]{\textbf{\ding{34}}} (h7);
      \draw (h3) -- (h4) ;
      \draw (h3) -- (h5) ;
      \draw (h2) -- (h6);

      \node (h11) at (14,0){$\bullet$};
      \node[blue] (h12) at (14,2){$\bullet$};
      \node[blue] (h13) at (12,4){$\bullet$};
      \node[blue] (h14) at (10,6){$\bullet$};
      \node[blue] (h15) at (14,6){$\bullet$};
      \node[red!90!black] (h16) at (16,4){$\bullet$};
      \draw (h11) -- (h12) ;
      \draw (h12) -- (h13) ;
      \draw (h12) -- node[red!90!black,
      font=\small,sloped,shift={(0.01,-0.075)},
      rotate=90]{\textbf{\ding{34}}} (h16);
      \draw (h13) -- (h14) ;
      \draw (h13) -- (h15) ;

      \node (hn1) at (24,0){$\bullet$};
      \node[blue] (hn2) at (22,2) {$\bullet$};
      \node[blue] (hn3) at (22,4) {$\bullet$};
      \node[blue] (hn4) at (20,6){$\bullet$};
      \node[blue] (hn5) at (22,6){$\bullet$};
      \draw (hn1) -- (hn2) ;
      \draw (hn2) -- (hn3) ;
      \draw (hn3) -- (hn4) ;
      \draw (hn3) -- (hn5) ;
      \node [blue](hn2b) at (24,2) {$\bullet$};
      \node[blue] (hn3b) at (24,4) {$\bullet$};
      \node [blue](hn4b) at (23.4,6){$\bullet$};
      \node [blue](hn5b) at (24.6,6){$\bullet$};
      \draw (hn1) -- (hn2b) ;
      \draw (hn2b) -- (hn3b) ;
      \draw (hn3b) -- (hn4b) ;
      \draw (hn3b) -- (hn5b) ;
      \node [blue](hn2c) at (26,2) {$\bullet$};
      \node [blue](hn3c) at (26,4) {$\bullet$};
      \node [blue](hn4c) at (26,6){$\bullet$};
      \node [red!90!black] (hn5c) at (28,6){$\bullet$};
      \draw (hn1) -- (hn2c) ;
      \draw (hn2c) -- (hn3c) ;
      \draw (hn3c) -- (hn4c) ;
      \draw (hn3c) -- node[red!90!black,
      font=\small,sloped,shift={(0.01,-0.075)},
      rotate=90]{\textbf{\ding{34}}} (hn5c) ;
    \end{tikzpicture}

    \caption{Three successive states of a hydra in a battle,
      at time $t=1,2,3$\label{life1}}
 \label{fig:round}
 \end{figure}

  \begin{figure}[h]
    \centering
    \begin{tikzpicture}[very thick, scale=0.2]
      \node (hn1) at (4,0){$\bullet$};
      \node[blue] (hn2) at (2,2) {$\bullet$};
      \node[blue] (hn3) at (2,4) {$\bullet$};
      \node[blue] (hn4) at (0,6){$\bullet$};
      \node[blue] (hn5) at (2,6){$\bullet$};
      \draw (hn1) -- (hn2) ;
      \draw (hn2) -- (hn3) ;
      \draw (hn3) -- (hn4) ;
      \draw (hn3) -- (hn5) ;
      \node [blue](hn2b) at (4,2) {$\bullet$};
      \node [blue](hn3b) at (4,4) {$\bullet$};
      \node [blue](hn4b) at (3.4,6){$\bullet$};
      \node [blue] (hn5b) at (4.6,6){$\bullet$};
      \draw (hn1) -- (hn2b) ;
      \draw (hn2b) -- (hn3b) ;
      \draw (hn3b) -- (hn4b) ;
      \draw (hn3b) -- (hn5b) ;
      \node [blue](hn2c) at (6,2) {$\bullet$};
      \node [blue](hn3c) at (6,4) {$\bullet$};
      \node [blue](hn4c) at (6,6){$\bullet$};
      \draw (hn1) -- (hn2c) ;
      \draw (hn2c) -- (hn3c) ;
      \draw (hn3c) -- (hn4c) ;
      \node [blue](hn3d) at (8,4) {$\bullet$};
      \draw (hn2c) -- (hn3d) ;
      \node [blue](hn4d) at (8,6) {$\bullet$};
      \draw (hn3d) -- (hn4d) ;

      \node [blue](hn3e) at (10,4) {$\bullet$};
      \draw (hn2c) -- (hn3e) ;
      \node [blue](hn4e) at (10,6) {$\bullet$};
      \draw (hn3e) -- (hn4e) ;

      \node [blue](hn3f) at (12,4) {$\bullet$};
      \draw (hn2c) -- (hn3f) ;
      \node [red!90!black](hn4f) at (12,6) {$\bullet$};
      \draw (hn3f) -- node[red!90!black,
      font=\small,sloped,shift={(0.01,-0.075)},
      rotate=90]{\textbf{\ding{34}}} (hn4f) ;

      \node at (16,0) {$\dots$} ;

      \node [black](hn4h) at (21,0) {$\bullet$};
      \node [red!90!black](hn4g) at (21,2) {$\bullet$};
      \draw (hn4h) -- node[red!90!black,
      font=\small,sloped,shift={(0.01,-0.075)},
      rotate=90]{\textbf{\ding{34}}} (hn4g) ;

      \node [black](hn4f) at (24,0) {$\bullet$};

    \end{tikzpicture}
    \caption{At $t=4,\dots,\textcolor{red}{x}-1,\textcolor{red}{x=?}$
      \label{life2}}
  \end{figure}

 Cf.  \textcolor{pauillac}{Pauillac 2006} and \textcolor{banyuls}{Banyuls 2018}. Today, let us focus on some combinatorial aspects (\emph{e.g.} number of rounds of a battle).
\end{frame}

@@

     \begin{block}{}
       \begin{itemize}
       \item In Coq, canonical sequences are defined by structural recursion on ordinal terms in Cantor normal form.

   {\color{mathcolor}
         $$\canonseq{\omega^{\omega^{\omega+1}+1}}{42}=
         \omega^{\omega^{\omega+1}}\times 42$$
         }

         \input{../../movies/snippets/Canon/canonExamplesb}
       \item Not the same definition as in K. \& S.'s article, but same abstract properties (monotony, limits, \dots). Sounds to be extentionally the same function.
       \end{itemize}
     \end{block}
\end{frame}

@@

          \label{fig:accessibility}
        \begin{tikzpicture}[very thick, scale=0.25]
\node (alpha) at (0,0) {$\alpha$};
    \node (beta) at (35, 0){\textcolor{red}{$\beta$}};
   \node (zero) at (40, 0){$0$};

@@

\only<2>{{\color{lookcolor}OK for large enough $i$ ! (K. \& S., 1981)}}
\vspace{6pt}

% \pause
\only<1>{Goal: reach \textcolor{red}{$\beta$} from $\alpha$.}
\only<2>{\textbf{Remark:} The proof is not trivial. The deterministic red path could miss $\beta$!}

@@

  {\small
\textbf{Hardy Hierarchy (variant)}

@@

 {\color{mathcolor}
\begin{align}
  H'_0(i) & = i\\
  H'_{\beta+1}(i) &=H'_\beta(i+1) \\
  H'_\alpha(i) &= H'_{(\canonseq{\alpha}{i+1})}(i)  \quad\textit{($\alpha$ limit ordinal)} 
\end{align}}

\textbf{Wainer hierarchy (variant)}

{\color{mathcolor}
\begin{align}
F_0(i)& =i+1\\
F_{\beta+1}(i)&= (F_\beta)^{(i+1)}(i)\\
F_\alpha(i) &= F_{(\canonseq{\alpha}{i})} (i) \quad \textit{($\alpha$ limit ordinal)}
\end{align}}}
\end{block}
\end{frame}

@@

        \begin{align*}
        H'_{\omega^\omega}(k)=\,& \textbf{let}\,F\;f\;i := f^{(i+1)}(i)\,
                                  \textbf{in}\, F^{(k)}(\lambda k.\,2k+1)\,k\\
      H'_{\omega^3}(k) \geq\, & \textit{hyper\_exp2}(k+1) \\
      H'_{\omega^\alpha}(k)\geq\,& F_\alpha(k)\; (k>0)\\
    \end{align*}
       }%mathcolor
  \end{block}

@@

\begin{frame}[label={fgftheorems}]
\label{thms-fastgrowing}
  \begin{block}{Theorems}
    \begin{itemize}
    \item \textcolor{mathcolor}{$F_\alpha$} is primitive recursive iff \textcolor{mathcolor}{$\alpha<\omega$}.
     \item  
    The function which computes the length of hydra battles
    for \textcolor{mathcolor}{$\alpha$} (according to the initial index) is not primitive recursive
    (for $\textcolor{mathcolor}{\alpha\geq \omega^\omega}$).
    \end{itemize}

@@

\begin{frame}
  \frametitle{}
  \begin{scriptsize}
    \inputsnippets{./SchemePrimRecInda}
  \end{scriptsize}
\end{frame}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Uncomment if there's time enough

% \section{Kirby \& Paris' hydras}

% \begin{frame}
%   \frametitle{Kirby \& Paris' hydras}
%  \begin{figure}[h]
%     \centering
%     \begin{tikzpicture}[very thick, scale=0.20]
%       \node (h1) at (4,0){$\bullet$};
%       \node[blue] (h2) at (4,2){$\bullet$};
%       \node[blue] (h3) at (2,4){$\bullet$};
%       \node[blue] (h4) at (0,6){$\bullet$};
%       \node[red!90!black] (h7) at (6,2){$\bullet$};
%       \node[blue] (h5) at (4,6){$\bullet$};
%       \node[blue] (h6) at (6,4){$\bullet$};
%       \draw (h1) -- (h2) ;
%       \draw (h2) -- (h3) ;
%       \draw (h1) -- node[red!90!black,font=\small,sloped,
%       shift={(0.01,-0.075)},rotate=90]{\textbf{\ding{34}}} (h7);
%       \draw (h3) -- (h4) ;
%       \draw (h3) -- (h5) ;
%       \draw (h2) -- (h6);

%       \node (h11) at (14,0){$\bullet$};
%       \node[blue] (h12) at (14,2){$\bullet$};
%       \node[blue] (h13) at (12,4){$\bullet$};
%       \node[blue] (h14) at (10,6){$\bullet$};
%       \node[blue] (h15) at (14,6){$\bullet$};
%       \node[red!90!black] (h16) at (16,4){$\bullet$};
%       \draw (h11) -- (h12) ;
%       \draw (h12) -- (h13) ;
%       \draw (h12) -- node[red!90!black,
%       font=\small,sloped,shift={(0.01,-0.075)},
%       rotate=90]{\textbf{\ding{34}}} (h16);
%       \draw (h13) -- (h14) ;
%       \draw (h13) -- (h15) ;

%       \node (hn1) at (24,0){$\bullet$};
%       \node[blue] (hn2) at (22,2) {$\bullet$};
%       \node[blue] (hn3) at (22,4) {$\bullet$};
%       \node[blue] (hn4) at (20,6){$\bullet$};
%       \node[blue] (hn5) at (22,6){$\bullet$};
%       \draw (hn1) -- (hn2) ;
%       \draw (hn2) -- (hn3) ;
%       \draw (hn3) -- (hn4) ;
%       \draw (hn3) -- (hn5) ;
%       \node [blue](hn2b) at (24,2) {$\bullet$};
%       \node[blue] (hn3b) at (24,4) {$\bullet$};
%       \node [blue](hn4b) at (23.4,6){$\bullet$};
%       \node [blue](hn5b) at (24.6,6){$\bullet$};
%       \draw (hn1) -- (hn2b) ;
%       \draw (hn2b) -- (hn3b) ;
%       \draw (hn3b) -- (hn4b) ;
%       \draw (hn3b) -- (hn5b) ;
%       \node [blue](hn2c) at (26,2) {$\bullet$};
%       \node [blue](hn3c) at (26,4) {$\bullet$};
%       \node [blue](hn4c) at (26,6){$\bullet$};
%       \node [red!90!black] (hn5c) at (28,6){$\bullet$};
%       \draw (hn1) -- (hn2c) ;
%       \draw (hn2c) -- (hn3c) ;
%       \draw (hn3c) -- (hn4c) ;
%       \draw (hn3c) -- node[red!90!black,
%       font=\small,sloped,shift={(0.01,-0.075)},
%       rotate=90]{\textbf{\ding{34}}} (hn5c) ;
%     \end{tikzpicture}

%     \caption{Three successive states of a hydra in a battle,
%       at time $t=1,2,3$\label{life1}}
%  \label{fig:round}
%  \end{figure}

%   \begin{figure}[h]
%     \centering
%     \begin{tikzpicture}[very thick, scale=0.2]
%       \node (hn1) at (4,0){$\bullet$};
%       \node[blue] (hn2) at (2,2) {$\bullet$};
%       \node[blue] (hn3) at (2,4) {$\bullet$};
%       \node[blue] (hn4) at (0,6){$\bullet$};
%       \node[blue] (hn5) at (2,6){$\bullet$};
%       \draw (hn1) -- (hn2) ;
%       \draw (hn2) -- (hn3) ;
%       \draw (hn3) -- (hn4) ;
%       \draw (hn3) -- (hn5) ;
%       \node [blue](hn2b) at (4,2) {$\bullet$};
%       \node [blue](hn3b) at (4,4) {$\bullet$};
%       \node [blue](hn4b) at (3.4,6){$\bullet$};
%       \node [blue] (hn5b) at (4.6,6){$\bullet$};
%       \draw (hn1) -- (hn2b) ;
%       \draw (hn2b) -- (hn3b) ;
%       \draw (hn3b) -- (hn4b) ;
%       \draw (hn3b) -- (hn5b) ;
%       \node [blue](hn2c) at (6,2) {$\bullet$};
%       \node [blue](hn3c) at (6,4) {$\bullet$};
%       \node [blue](hn4c) at (6,6){$\bullet$};
%       \draw (hn1) -- (hn2c) ;
%       \draw (hn2c) -- (hn3c) ;
%       \draw (hn3c) -- (hn4c) ;
%       \node [blue](hn3d) at (8,4) {$\bullet$};
%       \draw (hn2c) -- (hn3d) ;
%       \node [blue](hn4d) at (8,6) {$\bullet$};
%       \draw (hn3d) -- (hn4d) ;

%       \node [blue](hn3e) at (10,4) {$\bullet$};
%       \draw (hn2c) -- (hn3e) ;
%       \node [blue](hn4e) at (10,6) {$\bullet$};
%       \draw (hn3e) -- (hn4e) ;

%       \node [blue](hn3f) at (12,4) {$\bullet$};
%       \draw (hn2c) -- (hn3f) ;
%       \node [red!90!black](hn4f) at (12,6) {$\bullet$};
%       \draw (hn3f) -- node[red!90!black,
%       font=\small,sloped,shift={(0.01,-0.075)},
%       rotate=90]{\textbf{\ding{34}}} (hn4f) ;

%       \node at (16,0) {$\dots$} ;

%       \node [black](hn4h) at (21,0) {$\bullet$};
%       \node [red!90!black](hn4g) at (21,2) {$\bullet$};
%       \draw (hn4h) -- node[red!90!black,
%       font=\small,sloped,shift={(0.01,-0.075)},
%       rotate=90]{\textbf{\ding{34}}} (hn4g) ;

%       \node [black](hn4f) at (24,0) {$\bullet$};

%     \end{tikzpicture}
%     \caption{At $t=4,\dots,\textcolor{red}{x}-1,\textcolor{red}{x=?}$
%       \label{life2}}
%   \end{figure}

%  Cf.  \textcolor{pauillac}{Pauillac 2006} and \textcolor{banyuls}{Banyuls 2018}. Today, let us focus on some combinatorial aspects (\emph{e.g.} number of rounds of a battle).
% \end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \frametitle{Ordinals: a short reminder}

  \begin{figure}[h]
    \scalebox{0.6}{
  \centering
\fbox{\Large
  \begin{minipage}{0.4\linewidth}
  \begin{align*}
     &\textcolor{blue}{0},\,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,\ldots\\
&\textcolor{red}{\omega},\,\omega+1,\omega+2,\omega+3,\ldots\\
&\textcolor{red}{\omega\times 2},\,\omega\times 2+1,\ldots, \textcolor{red}{\omega\times 3},\,\omega\times 3+1,\ldots, \textcolor{red}{\omega\times 4},\ldots\\
&\textcolor{red}{\omega^2},\ldots, \textcolor{red}{\omega^2\times 42},\ldots,\textcolor{red}{\omega^3},\ldots, \textcolor{red}{\omega^4},\omega^4+1,\ldots\\
&\textcolor{red}{\omega^\omega},\ldots, \textcolor{red}{\omega^\omega+\omega^7\times 8},\ldots,\textcolor{red}{\omega^\omega\times 2},\omega^\omega\times 2+1, \ldots\\
&\textcolor{red}{\omega^{\omega^\omega}},\ldots, \textcolor{red}{\omega^{\omega^\omega}+\omega^\omega\times 42+ \omega^{55}+\omega}, \ldots, \textcolor{red}{\omega^{\omega^{\omega+1}}}, \omega^{\omega^{\omega+1}}+1,\dots\\
     & \textcolor{red!55}{\epsilon_0 (= \omega^{\epsilon_0)}},
       \textcolor{black!55}{\epsilon_0+1}, \textcolor{black!55}{\epsilon_0+2}, \textcolor{black!55}{\epsilon_0+3}, \ldots\\
& \textcolor{red!45}{\epsilon_1}, \ldots, \textcolor{red!45}{\epsilon_2}, \ldots, \textcolor{red!45}{\epsilon_\omega},\ldots \\
& \textcolor{red!40}{\Gamma_0}, \textcolor{black!40}{\Gamma_0+1}, \textcolor{black!40}{\Gamma_0+2}, \textcolor{black!40}{\Gamma_0+3},\ldots, \textcolor{red!40}{\Gamma_0+\omega}, \ldots\\
&\ldots
  \end{align*}   
  \end{minipage}}}

  %\caption{A short overview of the sequence of ordinal numbers}
  \label{fig:ordinal-sequence}
\end{figure}
\end{frame}

@@

    Canonical sequences are implemented by a structurally recursive function on terms in Cantor normal form.

\vspace{5pt}

    Example: $$\textcolor{mathcolor}{\canonseq{\omega^{\omega^{\omega+1}+1}}{42}}$$

      {\footnotesize
    \input{../../movies/snippets/Canon/canonExamplesb}
  }

   %   
    For efficiency's sake, our definition in \coq is sligthly different from Ketonen and Solovay's article (based on the decomposition on any ordinal \textcolor{mathcolor}{$\alpha$} into  \textcolor{mathcolor}{$\beta(\gamma+1)$}).

We \emph{believe} that both definitions are extentionally equivalent.

     %  \end{itemize}
     \end{block}
\end{frame}

@@

          \label{fig:accessibility}
          \begin{tikzpicture}[very thick, scale=0.25]
\node (epsilon0) at (-3,0) {\textcolor{black!60}{$(\epsilon_0>\;)$}};
\node (alpha) at (0,0) {$\alpha$};
    \node (beta) at (35, 0){\textcolor{red}{$\beta$}};
   \node (zero) at (40, 0){$0$};

@@

\only<2>{{\color{lookcolor}If $i$ is large enough, we are sure to reach $\beta$ (K. \& S., 1981)}}
\vspace{6pt}

% \pause
\only<1>{Can we  reach \textcolor{red}{$\beta$} from $\alpha$? Under which condition? }

@@

  {

@@

 {\color{mathcolor}
\begin{align}
  H'_0(i) & = i\\
  H'_{\beta+1}(i) &=H'_\beta(i+1) \\
  H'_\alpha(i) &= H'_{(\canonseq{\alpha}{i+1})}(i)  \quad\textit{($\alpha$ limit ordinal)} 
\end{align}}}

{$H'$ is a slight variant of the \emph{Hardy Hierarchy}. In \coq, we implement it with the \texttt{equations} plug-in.}
%\vspace{6pt}

% \textbf{Wainer hierarchy (variant)}

% {\color{mathcolor}
% \begin{align}
% F_0(i)& =i+1\\
% F_{\beta+1}(i)&= (F_\beta)^{(i+1)}(i)\\
% F_\alpha(i) &= F_{(\canonseq{\alpha}{i})} (i) \quad \textit{($\alpha$ limit ordinal)}
% \end{align}}}
\end{block}
\end{frame}

@@

        \begin{align*}
        H'_{\omega^\omega}(k)=\,& \textbf{let}\,F\;f\;i := f^{(i+1)}(i)\,
                                  \textbf{in}\, F^{(k)}(\lambda k.\,2k+1)\,k\\
      H'_{\omega^3}(k) \geq\, & \textit{hyper\_exp2}(k+1) \\
 %     H'_{\omega^\alpha}(k)\geq\,& F_\alpha(k)\; (k>0)\\
    \end{align*}
       }%mathcolor
  \end{block}

@@

\begin{frame}[label={fgftheorems}]
\label{thms-fastgrowing}
  \begin{block}{Theorem}
    % \begin{itemize}
    % \item \textcolor{mathcolor}{$F_\alpha$} is primitive recursive iff \textcolor{mathcolor}{$\alpha<\omega$}.
    %  \item  
    The function which computes the length of hydra battles
    for \textcolor{mathcolor}{$\alpha$} (according to the initial index) is not primitive recursive
    (for $\textcolor{mathcolor}{\alpha\geq \omega^\omega}$).
  %  \end{itemize}

@@

% \begin{frame}
%   \frametitle{}
%   \begin{scriptsize}
%     \inputsnippets{./SchemePrimRecInda}
%   \end{scriptsize}
% \end{frame}
